Milky way mass estimation using PCA-AGHQ

### 2023-07-04 src/astro/astro.Rmd
Fixes to astro example -- now understand where we are (needing a method for hyper marginals from PCA-AGHQ and EB) #46

---
title: "Milky way mass estimation using PCA-AGHQ"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background**

  **Task** 
---

We reanalyse the "estimating the mass of the Milky Way" example from @bilodeau2021stochastic.

@@

```{r}
ffa <- list(
  fn = function(x) -1 * ff$fn(x),
  gr = function(x) -1 * ff$gr(x),
  he = function(x) -1 * ff$he(x)
)

@@

```{r}
tm <- Sys.time()

@@

```{r}
C <- Matrix::forceSymmetric(solve(useropt$hessian))
eigenC <- eigen(C)
lambda <- eigenC$values

data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
) %>% ggplot(aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 3, y = 0.875, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()

k_pca <- 5
s <- 1 #' Could choose this based on the Scree plot
m <- length(useropt$mode)
levels <- c(rep(k_pca, s), rep(1, m - s))
pca_grid <- mvQuad::createNIGrid(dim = m, type = "GHe", level = levels)
mvQuad::rescale(pca_grid, m = useropt$mode, C = C, dec.type = 1)

@@

```{r}
#' Manual plots
parambounds <- list(
  Psi0 = c(1, 200),
  gamma = c(0.3, 0.7),
  alpha = c(3.0, 3.7),
  beta = c(-0.5, 1)
)

@@

Psi0_postplot <- Psi0pdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = Psi0, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = Psi0pcapdf,
    inherit.aes = FALSE,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = Psi0prior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(10, 60, by = 5)) +
  coord_cartesian(xlim = c(24, 43)) +
  theme_minimal()

gamma_postplot <- gammapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = gamma, y = after_stat(density)),
    bins = 100,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = gammapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = gammaprior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(0, 0.44, by = 0.02)) +
  coord_cartesian(xlim = c(0.3, 0.4)) +
  theme_minimal()

alpha_postplot <- alphapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = alpha, y = after_stat(density)),
    bins = 2000,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = alphapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = alphaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(3, 4, by = 0.02)) +
  coord_cartesian(xlim = c(3, 3.05)) +
  theme_minimal()

beta_postplot <- betapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = beta, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = betapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = betaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(-0.5, 1, by = 0.1)) +
  coord_cartesian(xlim = c(-0.3, 0.4)) +
  theme_minimal()

{Psi0_postplot + gamma_postplot} / {alpha_postplot + beta_postplot}
```

@@

```{r}
kstable <- function(quad) {
  aghqpostsamps_spline <- sample_marginal(astroquad, nrow(standata), interpolation = "spline")
  aghqpostsamps_poly <- sample_marginal(astroquad, nrow(standata), interpolation = "polynomial")

  suppressWarnings({
    kstable <- data.frame(
      Psi0 = ks.test(standata[[1]], aghqpostsamps_poly[[1]])$statistic,
      gamma = ks.test(standata[[2]], aghqpostsamps_spline[[2]])$statistic,
      alpha = ks.test(standata[[3]], aghqpostsamps_poly[[3]])$statistic,
      beta = ks.test(standata[[4]], aghqpostsamps_poly[[4]])$statistic
    )
  })

@@

kstable_astroebquad <- kstable(astroebquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "EB", k = k_eb, s = NA, total_points = k_eb)

kstable_astropcaquad <- kstable(astroquad) %>%
  pivot_longer(cols = everything()) %>%
  mutate(method = "AGHQ-PCA", k = k_pca, s = s, total_points = k^s)

bind_rows(kstable_astroquad, kstable_astroebquad, kstable_astropcaquad) %>%
  gt::gt()

bind_rows(kstable_astroquad, kstable_astroebquad, kstable_astropcaquad) %>%
  ggplot(aes(x = name, y = value, col = method)) +
    geom_jitter(height = 0, width = 0.05) +
    scale_y_continuous(limits = c(0, 0.05)) +
    scale_colour_manual(values = multi.utils::cbpalette()) +
    labs(x = "Parameter", y = "KS(method, NUTS)", col = "Method") +
    theme_minimal()
```

>>>>

---
title: "Milky way mass estimation using PCA-AGHQ"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Task** We reanalyse the "estimating the mass of the Milky Way" example from @bilodeau2021stochastic.
---

@@

```{r message=FALSE}
ffa <- list(
  fn = function(x) -1 * ff$fn(x),
  gr = function(x) -1 * ff$gr(x),
  he = function(x) -1 * ff$he(x)
)

@@

```{r message=FALSE}
tm <- Sys.time()

@@

```{r message=FALSE}
C <- Matrix::forceSymmetric(solve(useropt$hessian))
eigenC <- eigen(C)
lambda <- eigenC$values

data.frame(
  n = 1:length(lambda),
  tv = cumsum(lambda / sum(lambda))
) %>% ggplot(aes(x = n, y = tv)) +
  geom_point() +
  geom_hline(yintercept = 0.9, col = "grey", linetype = "dashed") +
  annotate("text", x = 3, y = 0.89, label = "90% of total variation explained", col = "grey") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "PCA dimensions included", y = "Total variation explained") +
  theme_minimal()

k_pca <- 5
s <- 1 #' Chosen based on the Scree plot
m <- length(useropt$mode)
levels <- c(rep(k_pca, s), rep(1, m - s))
pca_grid <- mvQuad::createNIGrid(dim = m, type = "GHe", level = levels)
mvQuad::rescale(pca_grid, m = useropt$mode, C = C, dec.type = 1)

@@

```{r warning=FALSE}
#' Manual plots
parambounds <- list(
  Psi0 = c(1, 200),
  gamma = c(0.3, 0.7),
  alpha = c(3.0, 3.7),
  beta = c(-0.5, 1)
)

@@

(Psi0_postplot <- Psi0pdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = Psi0, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = Psi0pcapdf,
    inherit.aes = FALSE,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = Psi0prior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(10, 60, by = 5)) +
  coord_cartesian(xlim = c(24, 43)) +
  theme_minimal())

(gamma_postplot <- gammapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = gamma, y = after_stat(density)),
    bins = 100,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = gammapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = gammaprior, linetype = "dashed") +
  labs(title = "", x = "", y = "Density") +
  scale_x_continuous(breaks = seq(0, 0.44, by = 0.02)) +
  coord_cartesian(xlim = c(0.3, 0.4)) +
  theme_minimal())

(alpha_postplot <- alphapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = alpha, y = after_stat(density)),
    bins = 2000,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = alphapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = alphaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(3, 4, by = 0.02)) +
  coord_cartesian(xlim = c(3, 3.05)) +
  theme_minimal())

(beta_postplot <- betapdf %>%
  ggplot(aes(x = transparam, y = pdf_transparam)) +
  geom_histogram(
    data = standata,
    mapping = aes(x = beta, y = after_stat(density)),
    bins = 50,
    colour = "white",
    fill = "grey"
  ) +
  geom_line() +
  geom_line(
    data = betapcapdf,
    mapping = aes(x = transparam, y = pdf_transparam),
    colour = "blue",
  ) +
  stat_function(fun = betaprior, linetype = "dashed") +
  labs(title = "",x = "",y = "Density") +
  scale_x_continuous(breaks = seq(-0.5, 1, by = 0.1)) +
  coord_cartesian(xlim = c(-0.3, 0.4)) +
  theme_minimal())
```

## Normalising constant

```{r}
data.frame(
  method = c("AGHQ", "PCA-AGHQ", "EB"),
  lognormconst = c(astroquad$normalized_posterior$lognormconst, astropcaquad$normalized_posterior$lognormconst, astroebquad$normalized_posterior$lognormconst)
) %>%
  gt::gt()
```

@@

```{r warning=FALSE}
kstable <- function(quad) {
  aghqpostsamps_spline <- sample_marginal(quad, nrow(standata), interpolation = "spline")
  aghqpostsamps_poly <- sample_marginal(quad, nrow(standata), interpolation = "polynomial")

  suppressWarnings({
    kstable <- data.frame(
      Psi0 = ks.test(standata[[1]], aghqpostsamps_poly[[1]])$statistic,
      gamma = ks.test(standata[[2]], aghqpostsamps_poly[[2]])$statistic,
      alpha = ks.test(standata[[3]], aghqpostsamps_poly[[3]])$statistic,
      beta = ks.test(standata[[4]], aghqpostsamps_poly[[4]])$statistic
    )
  })

@@

#' Not working (yet?)
# kstable_astropcaquad <- kstable(astropcaquad) %>%
#   pivot_longer(cols = everything()) %>%
#   mutate(method = "AGHQ-PCA", k = k_pca, s = s, total_points = k^s)

bind_rows(kstable_astroquad) %>%
  gt::gt()

bind_rows(kstable_astroquad) %>%
  ggplot(aes(x = name, y = value, col = method)) +
    geom_jitter(height = 0, width = 0.05) +
    scale_y_continuous(limits = c(0, 0.05)) +
    scale_colour_manual(values = multi.utils::cbpalette()) +
    labs(x = "Parameter", y = "KS(method, NUTS)", col = "Method") +
    theme_minimal()
```
